---
title: "guideline-reproducibility"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(palmerpenguins)

```

The goal of guideline-reproducibility is to introduce a method for package management that is helpful for reproducibility when collaborating in R. 

<BR>

## Motivating Example

Collaboration in R may become frustrating if differences in installed packages in collaborators' system libraries cause code to fail. It is not uncommon for code that runs without error in your R session, to encounter an error on a collaborator's machine due to package version differences. This can result in time-consuming debugging that takes time away from more important tasks.

<BR>

## renv 

The `renv` package can solve this issue by creating separate a project library for all packages that are installed and used in your project's code. `renv` stores metadata for each package in a lockfile that can be shared and will allow others to install the same package versions into their project library. Since the library is project-specific, you won't have to worry about `renv` messing with your global library of packages. 

<BR>

### Starting a renv project

`renv` is managed directly in the R console through just a few functions. 

When choosing to add `renv` to your project for the first time, you'll run

```{r getting-started, eval = FALSE}
renv::init()
```

This adds a [few files](https://rstudio.github.io/renv/articles/renv.html#getting-started) including the lockfile to your project. You'll never need to interact or edit these directly.


If you've already been working on your project for awhile, initiating the project may take longer than if you are starting a project from scratch because `renv` will install into your project specific library and record in the lockfile all packages it already knows you use in your project. 

<BR>

### Using a new package

Once initiating `renv`, you'll notice that packages you had installed for other projects, are no longer found in this project. If you end up needing a new package, you'll need to install it again, this time into your project library using either `install.packages()` or `renv::install()` like normal.

Let's say I want to use `dplyr` for the first time in the project. Running `install.pacakges` installs `dplyr` and all of its dependencies into my project library.

```{r, eval = FALSE}
install.packages("dplyr")
```

```
The following package(s) will be installed:
- dplyr      [1.1.4]
- fansi      [1.0.6]
- generics   [0.1.3]
- pillar     [1.9.0]
- pkgconfig  [2.0.3]
- tibble     [3.2.1]
- tidyselect [1.2.1]
- utf8       [1.2.4]
- withr      [3.0.1]
These packages will be installed into "./guideline-reproducibility/renv/library/R-4.2/x86_64-w64-mingw32".

Do you want to proceed? [Y/n]: Y

Installing packages --------------------------------------------------------
- Installing generics ...                       OK [linked from cache]
- Installing fansi ...                          OK [linked from cache]
- Installing utf8 ...                           OK [linked from cache]
- Installing pillar ...                         OK [linked from cache]
- Installing pkgconfig ...                      OK [linked from cache]
- Installing tibble ...                         OK [linked from cache]
- Installing withr ...                          OK [linked from cache]
- Installing tidyselect ...                     OK [linked from cache]
- Installing dplyr ...                          OK [linked from cache]
Successfully installed 9 packages in 0.18 seconds.

```

You may think the next step is recording the package and its dependencies in the lockfile. You do this through taking a `snapshot` of the project. 

```{r snapshot, eval = FALSE}
renv::snapshot()
```

```
- The lockfile is already up to date.
```

However, `renv` tells us the lockfile is already up to date and does not update the lockfile. This is because we actually aren't using `dplyr` anywhere in our code yet. This is nice because just having a package installed in your project library does not mean that it will get recorded in the lockfile which would later prompt your collaborator to install the package even though it's not used in the project.


If we add some code using `dplyr` and then call `renv::snapshot` again, `renv` will ask permission to update the lockfile by adding `dplyr` and its dependencies.

```{r dplyr-code, eval = FALSE}
library(dplyr)

dplyr_code <- penguins %>%
  filter(species == "Gentoo")
```

```{r snapshot2, eval = FALSE}
renv::snapshot()
```

```
The following package(s) will be updated in the lockfile:

CRAN -----------------------------------------------------------------------
- dplyr        [* -> 1.1.4]
- fansi        [* -> 1.0.6]
- tidyselect   [* -> 1.2.1]
- utf8         [* -> 1.2.4]
- withr        [* -> 3.0.1]

RSPM -----------------------------------------------------------------------
- generics     [* -> 0.1.3]
- pillar       [* -> 1.9.0]
- pkgconfig    [* -> 2.0.3]
- tibble       [* -> 3.2.1]

Do you want to proceed? [Y/n]: Y

- Lockfile written to "./guideline-reproducibility/renv.lock".
```

Here is a look into what the lockfile is recording about the package `dplyr`. You see the version number, where it was downloaded from, and all of the dependencies it relies on. Each of those dependencies also has a file in lockfile with this information. You can explore the whole lockfile by looking in `renv.lock` in the project directory.

```
"dplyr": {
      "Package": "dplyr",
      "Version": "1.1.4",
      "Source": "Repository",
      "Repository": "CRAN",
      "Requirements": [
        "R",
        "R6",
        "cli",
        "generics",
        "glue",
        "lifecycle",
        "magrittr",
        "methods",
        "pillar",
        "rlang",
        "tibble",
        "tidyselect",
        "utils",
        "vctrs"
      ],
      "Hash": "fedd9d00c2944ff00a0e2696ccf048ec"
    }
```

<BR>

If you are ever unsure if there is anything new to record with a snapshot, you can run

```{r status, eval = FALSE}
renv::status()
```

```
No issues found -- the project is in a consistent state.
```

<BR>

If we add in new code that relies on ggplot2, `renv::status` alerts us that there are installed and used packages that are not recorded in the lockfile.

```{r status2, eval = FALSE}
library(ggplot2)

ggplot(penguins,
        aes(x = species, y = body_mass_g, color = species)) + 
  geom_boxplot() + 
  geom_point(position = "jitter")

renv::status()
```

```
The following package(s) are in an inconsistent state:

 package      installed recorded used
 colorspace   y         n        y   
 farver       y         n        y   
 ggplot2      y         n        y   
 gtable       y         n        y   
 isoband      y         n        y   
 labeling     y         n        y   
 lattice      y         n        y   
 MASS         y         n        y   
 Matrix       y         n        y   
 mgcv         y         n        y   
 munsell      y         n        y   
 nlme         y         n        y   
 RColorBrewer y         n        y   
 scales       y         n        y   
 viridisLite  y         n        y   
```

The solution is to call `renv::snapshot` to update the lockfile again.

```{r snapshot3, eval = FALSE}
renv::snapshot()
```

```
The following package(s) will be updated in the lockfile:

CRAN -----------------------------------------------------------------------
- colorspace     [* -> 2.1-1]
- farver         [* -> 2.1.2]
- gtable         [* -> 0.3.5]
- isoband        [* -> 0.2.7]
- labeling       [* -> 0.4.3]
- lattice        [* -> 0.20-45]
- MASS           [* -> 7.3-58.2]
- Matrix         [* -> 1.5-3]
- mgcv           [* -> 1.8-42]
- munsell        [* -> 0.5.1]
- nlme           [* -> 3.1-162]
- viridisLite    [* -> 0.4.2]

RSPM -----------------------------------------------------------------------
- ggplot2        [* -> 3.5.1]
- RColorBrewer   [* -> 1.1-3]
- scales         [* -> 1.3.0]

Do you want to proceed? [Y/n]: Y

- Lockfile written to "./guideline-reproducibility/renv.lock".
```

<BR>

### Sharing with Collaborators

When your collaborator opens your shared project files on their own computer for the first time with `renv`, `renv` will be automatically installed, and they will be prompted to run

```{r restore, eval = FALSE}
renv::restore()
```

The function will download and install all packages of the proper versions recorded in the lockfile into the project-specific library. 

<BR>

### Try it yourself

Clone `guidelines-reproducibility` on your own computer and run `renv::restore()` to install the currently recorded packages into your project library. Then experiment with adding new packages to `packages.R` or write R scripts using new packages to this project. Explore how you can use `renv::status` and `renv::snapshot` to get the lockfile back up to date.



For more information on `renv`, see the following vignettes: 

- [Introduction to renv](https://rstudio.github.io/renv/articles/renv.html)

- [FAQs](https://rstudio.github.io/renv/articles/faq.html)
